Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.tex
After weird default latex layout occurred

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R}. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence. With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged \emph{enough} for valid inferences. We translate the results of the study into guidelines for applied researchers. \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}}

@@

\caption{Typical convergence versus pathological non-convergence. Depicted are are chain means of variable $j$ in each imputation ($m=5$; $T=10$). Please note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$.}\label{fig:non-conv}
\end{figure}

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. In \texttt{mice}, the default scalar summaries to inspect are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep{buur18}.

@@

As alternative, van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.

@@

When we look at panel B, we conclude something weird. The \(AC\)-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence scenarios (up-to \(t=5\)), while there is obvious trending in the \(\theta\)s of the latter. Moreover, the best convergence (as indicated by the lowest \(AC\)-value) is observed at \(t=2\) for both scenarios. However, if we look at the chain means of the non-convergence scenario, there should be signs of trending up-to iteration number seven. After consulting the documentation on \texttt{stats::acf()}, we conclude that this \(AC\) function is not suited for iterative imputation algorithms. The function is optimized for performance when \(t\geq50\) \citep{box15}, while the default number of iterations in iterative imputation is often much lower. Therefore, we compute \(AC\) manually, see the solid line. These \(AC\) values behave as expected from visual inspection. We will, therefore, only consider manually calculated \(AC\)-values as non-convergence diagnostic.

@@

The assumption inherent to the second set of simulation conditions, early stopping, is that terminating the imputation algorithm too early causes non-convergence. Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the initial values of the algorithm \citep[which are sampled randomly from the set of observed datapoints;][]{buur18}. As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state). We expect that we can induce non-convergence by early stopping in at least those conditions where \(T\) is smaller than the default number of iterations in \texttt{mice}, \(T=5\) \citep{mice}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high proportion of incomplete cases \(p_{\rm inc}\) and a low number of iterations \(T\) will more often result in biased, invalid estimates of the quantities of scientific interest \(Q\).
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where any \(\bar{Q}\) is \emph{not} an unbiased and confidence-valid estimate of \(Q\).
\item
  We hypothesize that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\), \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too stringent for iterative imputation applications. In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}. We, therefore, suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates. In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
\item
  We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence. After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
\item
  We further hypothesize that multivariate \(\theta\)s are better at detecting non-convergence than univariate \(\theta\)s.
\end{enumerate}

@@

In each simulation repetition, we \emph{ampute} the complete dataset conform the `incompleteness' conditions. The proportion of incomplete cases \(p_{\rm inc}\) varies between 5\% and 95\%, in incremental steps (\(p_{\rm inc}\) = .05, .25, .50, .75, .95). We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep[MCAR;][]{rubin87}. I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on \(p_{\rm inc}\). We use the `mice' package \citep[function \texttt{mice::ampute()};][]{mice} to obtain an incomplete dataset \{\(y_{\rm obs}, y_{\rm mis}\)\} for each of the five `incompleteness' conditions.

We impute the missing values in \{\(y_{\rm obs}, y_{\rm mis}\)\} according to the `early stopping' conditions. That is, we vary the number of iterations in the imputation algorithm between one and one hundred (\(T = 1,2,.,100\)). All imputation procedures are performed using \texttt{mice} \citep[function \texttt{mice()};][]{mice}, with Bayesian linear regression imputation, and five imputation chains (\(m=5\)). This results in five sets of imputations for each of the one hundred `early stopping' conditions, \(y_{\rm imp, \ell}\).

@@

Our results show the performance of the MICE algorithm under different conditions of `incompleteness' and `early stopping'. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands (\(Q=\mu\) and \(Q=\sigma\)) we observe the largest bias in the outcome variable \(Y\). And for the regression coefficients (\(Q=\beta\)) the bias is most pronounced in \(\beta_1\) (the effect of \(X_1\) on \(Y\)). Since there is just one estimate for \(Q=r^2\) to consider, we \textbf{end up with} \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\). In the figures, we present all `incompleteness' conditions, but only `early stopping' conditions where \(1 \leq T\leq50\). That is, the results are more or less stable for conditions where \(T \geq 30\). Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}.

@@

\noindent Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity. Under low to moderate missingness, it does not seems to matter on which \(\theta\) the non-convergence diagnostics are applied. If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases \(p_{\rm inc}\) and a low number of iterations \(T\). However, this only holds for multivariate scientific estimands. Univariate \(Q\)s are not substantially impacted by early stopping.
\item
  Our results also support the hyothesis that \(\widehat{R}\) and \(AC\) would correctly identify signs of non-convergence. We observed higher \(\widehat{R}\) and \(AC\)-values in conditions where \(\bar{Q}\)s were biased, invalid estimates of \(Q\)s.
\item
  We hypothesized that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) would be too stringent for iterative imputation applications. This hypothesis is only partially supported. \(\widehat{R}\) indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations. This roughly corresponds to the threshold of 1.2. The more stringent threshold 1.1 seems plausible compared to the decrease in \(\widehat{R}\) values across iterations. The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
\item
  The results of this simulation study do not support the hypothesis that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence. We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations. In this study, however, we observed \(AC\)-values as high as theoretically possible (\(AC=1\)). The second part of the hypothesis is not refuted: the \(AC\)-values did decrease quickly as a function of \(T\).
\item
  Our final hypothesis was that multivariate \(\theta\)s would be better at detecting non-convergence than univariate \(\theta\)s. Multivariate \(\theta\)s did indeed show superior performance under certain conditions (e.g., multivariate \(Q\)s and extreme missingness proportions).
\end{enumerate}

>>>>

In this paper we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R}. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence. With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged \emph{enough} for valid inferences. We translate the results of the study into guidelines for applied researchers. \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}}

@@

\caption{Typical convergence versus pathological non-convergence. Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$). Please note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$.}\label{fig:non-conv}
\end{figure}

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep{buur18}.

@@

As alternative, van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in the \texttt{mice} package \citep{mice}.

@@

When we look at panel B, we conclude something weird. The \(AC\)-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence scenarios (up-to \(t=5\)), while there is obvious trending in the \(\theta\)s of the latter. Moreover, the best convergence (as indicated by the lowest \(AC\)-value) is observed at \(t=2\) for both scenarios. However, if we look at the chain means of the non-convergence scenario, there should be signs of trending up-to iteration number seven. After consulting the documentation on \texttt{stats::acf()}, we conclude that this \(AC\) function is not suited for iterative imputation algorithms. The function is optimized for performance when \(t\geq50\) \citep{box15}, while the default number of iterations in iterative imputation is often much lower. Therefore, we compute \(AC\) manually, see the solid line. These \(AC\)-values behave as expected from visual inspection. We will, therefore, only consider manually calculated \(AC\)-values as non-convergence diagnostic.

@@

The assumption inherent to the early stopping conditions is that terminating the imputation algorithm too early causes non-convergence. Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the initial values of the algorithm \citep[which are sampled randomly from the set of observed datapoints;][]{buur18}. As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state). We expect that we can induce non-convergence by early stopping in at least those conditions where \(T\) is smaller than the default number of iterations in \texttt{mice}, \(T=5\) \citep{mice}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high proportion of incomplete cases \(p_{\rm inc}\) and a low number of iterations \(T\) will more often result in biased, invalid estimates of the quantities of scientific interest \(Q\).
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where any \(\bar{Q}\) is \emph{not} an unbiased and confidence-valid estimate of \(Q\).
\item
  We hypothesize that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\), \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too stringent for iterative imputation applications. In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}. We, therefore, suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates. In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
\item
  We expect that high \(AC\)-values are implausible in iterative imputation algorithms with typical convergence. After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
\item
  We further hypothesize that multivariate \(\theta\)s are better at detecting non-convergence than univariate \(\theta\)s.
\end{enumerate}

@@

In each simulation repetition, we \emph{ampute} the complete dataset conform the incompleteness conditions. The proportion of incomplete cases \(p_{\rm inc}\) varies between 5\% and 95\%, in incremental steps (\(p_{\rm inc}\) = .05, .25, .50, .75, .95). We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep[MCAR;][]{rubin87}. I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on \(p_{\rm inc}\). We use the `mice' package \citep[function \texttt{mice::ampute()};][]{mice} to obtain an incomplete dataset \{\(y_{\rm obs}, y_{\rm mis}\)\} for each of the five incompleteness conditions.

We impute the missing values in \{\(y_{\rm obs}, y_{\rm mis}\)\} according to the early stopping conditions. That is, we vary the number of iterations in the imputation algorithm between one and one hundred (\(T = 1,2,.,100\)). All imputation procedures are performed using \texttt{mice} \citep[function \texttt{mice()};][]{mice}, with Bayesian linear regression imputation, and five imputation chains (\(m=5\)). This results in five sets of imputations for each of the one hundred early stopping conditions, \(y_{\rm imp, \ell}\).

@@

Our results show the performance of the MICE algorithm under different conditions of incompleteness and early stopping. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands (\(Q=\mu\) and \(Q=\sigma\)) we observe the largest bias in the outcome variable \(Y\). And for the regression coefficients (\(Q=\beta\)) the bias is most pronounced in \(\beta_1\) (the effect of \(X_1\) on \(Y\)). Since there is just one estimate for \(Q=r^2\) to consider, we \textbf{end up with} \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\). In the figures, we present all incompleteness conditions, but only early stopping conditions where \(1 \leq T\leq50\). That is, the results are more or less stable for conditions where \(T \geq 30\). Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}.

@@

\noindent Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity. Under low to moderate missingness, it does not seem to matter on which \(\theta\) the non-convergence diagnostics are applied. If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases \(p_{\rm inc}\) and a low number of iterations \(T\). However, this only holds for multivariate scientific estimands. Univariate \(Q\)s are not substantially impacted by early stopping.
\item
  Our results also support the hyothesis that \(\widehat{R}\) and \(AC\) would correctly identify signs of non-convergence. We observed higher \(\widehat{R}\) and \(AC\)-values in conditions where \(\bar{Q}\)s were biased, invalid estimates of \(Q\)s.
\item
  We hypothesized that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) would be too stringent for iterative imputation applications. This hypothesis is only partially supported. \(\widehat{R}\) indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations. This roughly corresponds to the threshold of 1.2. The more stringent threshold 1.1 seems plausible compared to the decrease in \(\widehat{R}\)-values across iterations. The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
\item
  The results of this simulation study do not support the hypothesis that high \(AC\)-values are implausible in iterative imputation algorithms with typical convergence. We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations. In this study, however, we observed \(AC\)-values as high as theoretically possible (\(AC=1\)). The second part of the hypothesis is not refuted: the \(AC\)-values did decrease quickly as a function of \(T\).
\item
  Our final hypothesis was that multivariate \(\theta\)s would be better at detecting non-convergence than univariate \(\theta\)s. Multivariate \(\theta\)s did indeed show superior performance under certain conditions (e.g., multivariate \(Q\)s and extreme missingness proportions).
\end{enumerate}
